---
title: "Difference-in-Differences with Double/Debiased Machine Learning"
author: "Thomas Wiemann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Difference-in-Differences with Double/Debiased Machine Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_ddml <- requireNamespace("ddml", quietly = TRUE)
```

## Introduction

This vignette illustrates implementation of a staggered difference-in-differences (DiD) estimator using double/debiased machine learning (DML). Instead of relying on ad-hoc parametric nuisance estimators, this DiD-DML estimator leverages machine learning to flexibly control for covariates.

The key implementation approach is simple: the `att_gt` function allows users to pass a custom estimation method via the `est_method` argument.  Any additional arguments passed to `att_gt(...)` are forwarded to the custom function via `...`, making it straightforward to pass learner configurations, cross-fitting parameters, and other options.

For a detailed discussion of the relevant asymptotic theory for DiD-DML estimators see Chang (2020). Ahrens et al. (2025) provide an introduction to DML along with an empirical example in the staggered DiD setting.


## Baseline: The Default Linear Estimator

Consider the data of Callaway and Sant'Anna (2021) on county-level teen employment rates from 2003--2007.  We are interested in the effect of treatment on the log-employment rate `lemp` and assume that parallel trends holds conditional on county population (`lpop`).

```{r data}
library(did)
set.seed(588239)
data(mpdta)
head(mpdta)
```

By default, `att_gt` controls *linearly* for covariates (logistic propensity score, linear outcome regression):

```{r default}
attgt_lm <- att_gt(yname = "lemp",
                   gname = "first.treat",
                   idname = "countyreal",
                   tname = "year",
                   xformla = ~lpop,
                   data = mpdta)
summary(attgt_lm)
```

```{r default_dyn, fig.width=8, fig.height=5, fig.align='center', out.width="90%", dpi = 200}
dyn_lm <- aggte(attgt_lm, type = "dynamic")
summary(dyn_lm)
ggdid(dyn_lm, ylim = c(-.4, .4))
```

Without additional *parametric* functional form assumptions on the nuisance functions, it is *not* guaranteed that the above estimator returns a convex combination of causal effects.  An important robustness check is therefore to verify whether the results hold up when using *nonparametric* nuisance estimators.


## A `ddml`-based Wrapper for `att_gt`

The [**ddml**](https://thomaswiemann.com/ddml/) package facilitates the use of a large set of machine learning nuisance estimators in a DML framework, including simultaneous consideration of multiple estimators via (short-)stacking.

To use **ddml** estimators with `att_gt`, we pass a custom function via `est_method`.  It is useful to construct this in two steps:

1. A general wrapper that adapts `ddml_att` to the interface expected by `att_gt`
2. Specific wrappers that fix learner configurations

Here is the general wrapper:

```{r ddml_wrapper, eval = has_ddml}
library(ddml)

ddml_did_method <- function(y1, y0, D, covariates, i.weights, inffunc, ...) {
  delta_y <- y1 - y0
  att_fit <- ddml_att(y = delta_y, D = D, X = covariates, ...)
  inf.func <- att_fit$psi_b + att_fit$att * att_fit$psi_a
  list(ATT = att_fit$att, att.inf.func = inf.func)
}
```

Note how `...` flows from `att_gt()` through to `ddml_att()`, allowing us to pass any arguments that `ddml_att` accepts.  We now construct three learner-specific wrappers for our robustness analysis.


## Robustness Check: Nonparametric Nuisance Estimators

We consider three nonparametric specifications to compare against the default linear estimator:

1. **Gradient tree boosting** via `mdl_xgboost` (300 rounds, depth 1)
2. **Random forests** via `mdl_ranger` (1000 trees, minimum node size 10)
3. **Short-stacking** combining the parametric specification with gradient tree boosting and random forests

All specifications use 10-fold cross-fitting.


### Gradient Tree Boosting

```{r xgboost_wrapper, eval = has_ddml}
my_did_xgboost <- function(y1, y0, D, covariates, i.weights, inffunc, ...) {
  learners <- list(what = mdl_xgboost,
                   args = list(nround = 300,
                               params = list(max_depth = 1)))
  ddml_did_method(y1, y0, D, covariates,
                  learners = learners,
                  learners_DX = learners,
                  sample_folds = 10,
                  silent = TRUE)
}
```

```{r xgboost_run, eval = has_ddml}
attgt_xgboost <- att_gt(yname = "lemp",
                         gname = "first.treat",
                         idname = "countyreal",
                         tname = "year",
                         xformla = ~lpop,
                         data = mpdta,
                         est_method = my_did_xgboost)
```


### Random Forests

```{r ranger_wrapper, eval = has_ddml}
my_did_ranger <- function(y1, y0, D, covariates, i.weights, inffunc, ...) {
  learners <- list(what = mdl_ranger,
                   args = list(num.trees = 1000,
                               min.node.size = 10))
  ddml_did_method(y1, y0, D, covariates,
                  learners = learners,
                  learners_DX = learners,
                  sample_folds = 10,
                  silent = TRUE)
}
```

```{r ranger_run, eval = has_ddml}
attgt_ranger <- att_gt(yname = "lemp",
                        gname = "first.treat",
                        idname = "countyreal",
                        tname = "year",
                        xformla = ~lpop,
                        data = mpdta,
                        est_method = my_did_ranger)
```


### Short-Stacking

Instead of committing to a single learner, we can simultaneously consider multiple learners and optimally combine them via non-negative least squares (NNLS).  Crucially, the stacking ensemble includes the linear/logistic specifications used by **did**'s default estimator, so machine learners are only selected if they improve out-of-sample fit.

To enable inspection of the stacking diagnostics, we store the `ddml_att` fit objects for each group-time pair:

```{r stacking_wrapper, eval = has_ddml}
# Store ddml_att fit objects for diagnostics
ddml_fits <- list()

my_did_stacking <- function(y1, y0, D, covariates, i.weights, inffunc, ...) {
  learners <- list(list(fun = ols),
                   list(fun = mdl_xgboost,
                        args = list(nround = 100,
                                    params = list(max_depth = 1))),
                   list(fun = mdl_xgboost,
                        args = list(nround = 300,
                                    params = list(max_depth = 1))),
                   list(fun = mdl_ranger,
                        args = list(num.trees = 1000,
                                    min.node.size = 10)),
                   list(fun = mdl_ranger,
                        args = list(num.trees = 1000,
                                    min.node.size = 100)))
  learners_DX <- list(list(fun = mdl_glm),
                      list(fun = mdl_xgboost,
                           args = list(nround = 100,
                                       params = list(max_depth = 1))),
                      list(fun = mdl_xgboost,
                           args = list(nround = 300,
                                       params = list(max_depth = 1))),
                      list(fun = mdl_ranger,
                           args = list(num.trees = 1000,
                                       min.node.size = 10)),
                      list(fun = mdl_ranger,
                           args = list(num.trees = 1000,
                                       min.node.size = 100)))
  delta_y <- y1 - y0
  att_fit <- ddml_att(y = delta_y, D = D, X = covariates,
                      learners = learners,
                      learners_DX = learners_DX,
                      sample_folds = 10,
                      ensemble_type = "nnls",
                      shortstack = TRUE,
                      silent = TRUE)
  # Store the full fit for diagnostics
  ddml_fits[[length(ddml_fits) + 1]] <<- att_fit
  inf.func <- att_fit$psi_b + att_fit$att * att_fit$psi_a
  list(ATT = att_fit$att, att.inf.func = inf.func)
}
```

```{r stacking_run, eval = has_ddml}
attgt_stacking <- att_gt(yname = "lemp",
                          gname = "first.treat",
                          idname = "countyreal",
                          tname = "year",
                          xformla = ~lpop,
                          data = mpdta,
                          est_method = my_did_stacking)
```


## Comparing Results

We can compare the group-time average treatment effects across all four estimators:

```{r comparison, eval = has_ddml}
comparison <- data.frame(
  group = attgt_lm$group,
  time = attgt_lm$t,
  linear = round(attgt_lm$att, 4),
  xgboost = round(attgt_xgboost$att, 4),
  ranger = round(attgt_ranger$att, 4),
  stacking = round(attgt_stacking$att, 4)
)
print(comparison)
```

The dynamic treatment effects tell a similar story:

```{r comparison_dyn, eval = has_ddml}
dyn_xgboost <- aggte(attgt_xgboost, type = "dynamic")
dyn_ranger <- aggte(attgt_ranger, type = "dynamic")
dyn_stacking <- aggte(attgt_stacking, type = "dynamic")

dyn_comparison <- data.frame(
  event_time = dyn_lm$egt,
  linear = round(dyn_lm$att.egt, 4),
  xgboost = round(dyn_xgboost$att.egt, 4),
  ranger = round(dyn_ranger$att.egt, 4),
  stacking = round(dyn_stacking$att.egt, 4)
)
print(dyn_comparison)
```

```{r stacking_dyn_plot, eval = has_ddml, fig.width=8, fig.height=5, fig.align='center', out.width="90%", dpi = 200}
ggdid(dyn_stacking, ylim = c(-.4, .4))
```


## Learner Diagnostics

The stacking wrapper stored the full `ddml_att` fit objects, including the short-stacking weights for each learner.  We can summarize these across all group-time pairs to understand which learners contribute to the final estimates.

```{r diagnostics, eval = has_ddml}
learner_names <- c("OLS/Logit", "XGBoost (nround=100)", "XGBoost (nround=300)",
                   "Ranger (nodes=10)", "Ranger (nodes=100)")
n_learners <- length(learner_names)
n_fits <- length(ddml_fits)

# Extract stacking weights across all (g,t) pairs
w_y_mat <- matrix(NA, nrow = n_fits, ncol = n_learners)
w_d_mat <- matrix(NA, nrow = n_fits, ncol = n_learners)
for (i in seq_len(n_fits)) {
  w_y_mat[i, ] <- ddml_fits[[i]]$weights[[1]][, 1]
  w_d_mat[i, ] <- ddml_fits[[i]]$weights[[2]][, 1]
}
colnames(w_y_mat) <- learner_names
colnames(w_d_mat) <- learner_names

# Compute weight summary statistics across (g,t) pairs
summarize_weights <- function(w_mat) {
  data.frame(min  = apply(w_mat, 2, min),
             mean = apply(w_mat, 2, mean),
             max  = apply(w_mat, 2, max))
}

cat("Outcome reduced form---stacking weights:\n")
print(round(summarize_weights(w_y_mat), 4))

cat("\nPropensity score---stacking weights:\n")
print(round(summarize_weights(w_d_mat), 4))
```

In this application with a single covariate, we expect the parametric learners (OLS/logit) to receive most of the stacking weight---reassuring us that the default linear estimator is adequate here.

In applications with many control variables or complex interactions, one would typically observe more weight on the nonparametric learners. See Ahrens et al. (2025) for further discussion.


## References

Ahrens A, Chernozhukov V, Hansen C, Kozbur D, Schaffer M, Wiemann T (2025). "An Introduction to Double/Debiased Machine Learning." Journal of Economic Literature, forthcoming.

Callaway B, Sant'Anna P (2021). "Difference-in-differences with multiple time periods." Journal of Econometrics, 200--230.

Chang NC (2020). "Double/debiased machine learning for difference-in-difference models." The Econometrics Journal, 177--191.
