---
title: "Difference-in-Differences with Double/Debiased Machine Learning"
author: "Thomas Wiemann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Difference-in-Differences with Double/Debiased Machine Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_ddml <- requireNamespace("ddml", quietly = TRUE)
```

## Introduction

This vignette illustrates implementation of a staggered difference-in-differences (DiD) estimator using double/debiased machine learning (DML). Instead of relying on ad-hoc parametric nuisance estimators, this DiD-DML estimator leverages machine learning to flexibly control for covariates.

The key implementation approach is simple: the `att_gt` function allows users to pass a custom estimation method via the `est_method` argument.  Any additional arguments passed to `att_gt(...)` are forwarded to the custom function via `...`, making it straightforward to pass learner configurations, cross-fitting parameters, and other options.

The vignette also demonstrates two features of `att_gt` that enable clean integration with `ddml`:

- **`est_method_vars`**: passes additional variables (e.g., pre-computed fold assignments) from the original data to the custom estimation function, enabling consistent cross-fitting across group-time cells.
- **`extra_gt`**: captures additional outputs from the custom estimation function for each group-time cell, which we use to store the full `ddml` fit objects for diagnostics and re-estimation.

For a detailed discussion of the relevant asymptotic theory for DiD-DML estimators see Chang (2020). Ahrens et al. (2025) provide an introduction to DML along with an empirical example in the staggered DiD setting.


## Baseline: The Default Linear Estimator

Consider the data of Callaway and Sant'Anna (2021) on county-level teen employment rates from 2003--2007.  We are interested in the effect of treatment on the log-employment rate `lemp` and assume that parallel trends holds conditional on county population (`lpop`).

```{r data}
library(did)
set.seed(588239)
data(mpdta)
head(mpdta)
```

By default, `att_gt` controls *linearly* for covariates (logistic propensity score, linear outcome regression):

```{r default}
attgt_lm <- att_gt(yname = "lemp",
                   gname = "first.treat",
                   idname = "countyreal",
                   tname = "year",
                   xformla = ~lpop,
                   data = mpdta)
summary(attgt_lm)
```

```{r default_dyn, fig.width=8, fig.height=5, fig.align='center', out.width="90%", dpi = 200}
dyn_lm <- aggte(attgt_lm, type = "dynamic")
summary(dyn_lm)
ggdid(dyn_lm, ylim = c(-.4, .4))
```

Without additional *parametric* functional form assumptions on the nuisance functions, it is *not* guaranteed that the above estimator returns a convex combination of causal effects.  An important robustness check is therefore to verify whether the results hold up when using *nonparametric* nuisance estimators.


## A `ddml`-based Wrapper for `att_gt`

The [**ddml**](https://thomaswiemann.com/ddml/) package facilitates the use of a large set of machine learning nuisance estimators in a DML framework, including simultaneous consideration of multiple estimators via (short-)stacking.

To use **ddml** estimators with `att_gt`, we pass a custom function via `est_method`.  The wrapper adapts `ddml_att` to the interface expected by `att_gt`:

```{r ddml_wrapper, eval = has_ddml}
library(ddml)

ddml_did_method <- function(y1, y0, D, covariates, i.weights, inffunc,
                            data = NULL, ...) {
  delta_y <- y1 - y0

  # Build ddml-compatible splits from the pre-assigned fold_id column.
  # We supply the overall fold indices via D_X; ddml auto-derives the
  # D-stratified subsamples by intersecting with D==0 and D==1 indices.
  fold_id <- data[["fold_id"]]
  K <- max(fold_id)
  subsamples <- lapply(seq_len(K), function(k) which(fold_id == k))
  splits <- list(
    D_X = list(subsamples = subsamples, cv_subsamples = NULL),
    y_X_D0 = list(subsamples = NULL, cv_subsamples = NULL),
    y_X_D1 = list(subsamples = NULL, cv_subsamples = NULL)
  )

  att_fit <- do.call(ddml_att, c(list(y = delta_y, D = D, X = covariates,
                                      splits = splits), list(...)))

  # ddml stores the evaluated Neyman-orthogonal score in a 3D array
  # scores[n_obs, 1, n_ensemble_types]; scores[, 1, 1] extracts the
  # n-vector of influence function values for the first ensemble type
  list(ATT = coef(att_fit),
       att.inf.func = att_fit$scores[, 1, 1],
       ddml_fit = att_fit)
}
```

Two implementation notes:

1. **Fold consistency via `est_method_vars`:**  The wrapper reads `fold_id` from the `data` argument (passed via `est_method_vars`) and constructs `ddml`-compatible `splits`.  This ensures all group-time cells and all learners use the same unit-level fold assignments.  Only the overall fold indices (`D_X$subsamples`) need to be supplied; `ddml` automatically derives the D-stratified subsamples for the outcome equations by intersecting with the $D=0$ and $D=1$ indices.

2. **Extra outputs via `extra_gt`:** Returning the full `ddml_att` fit as `ddml_fit` makes it available on the output object via `result$extra_gt` for per-cell diagnostics and re-estimation; see below.


## Robustness Check: Nonparametric Nuisance Estimators

We consider three nonparametric specifications to compare against the default linear estimator:

1. **Gradient tree boosting (depth 1)** via `mdl_xgboost` (300 rounds, max depth 1)
2. **Gradient tree boosting (depth 2)** via `mdl_xgboost` (300 rounds, max depth 2)
3. **Short-stacking** combining the parametric specification with multiple gradient tree boosting configurations

All specifications use 20-fold cross-fitting with consistent fold assignments.  We use a relatively high number of folds here so that the training sets remain large even for the smaller group-time cells.

First, we pre-assign fold IDs at the unit level:

```{r fold_setup, eval = has_ddml}
K <- 20
mpdta$fold_id <- sample(seq_len(K), nrow(mpdta), replace = TRUE)
```

Of course more advanced subsampling schemes are also possible here, including stratification across treatment cohorts.

### Gradient Tree Boosting (depth 1)

```{r xgboost_run, eval = has_ddml, warning = FALSE}
attgt_xgboost <- att_gt(
  yname = "lemp",
  gname = "first.treat",
  idname = "countyreal",
  tname = "year",
  xformla = ~lpop,
  data = mpdta,
  est_method = ddml_did_method,
  est_method_vars = "fold_id",
  # --- forwarded to ddml_att via ... ---
  learners = list(what = mdl_xgboost,
                  args = list(nround = 300,
                              max_depth = 1)),
  sample_folds = K,
  silent = TRUE)
```


### Gradient Tree Boosting (depth 2)

```{r xgboost2_run, eval = has_ddml, warning = FALSE}
attgt_xgboost2 <- att_gt(
  yname = "lemp",
  gname = "first.treat",
  idname = "countyreal",
  tname = "year",
  xformla = ~lpop,
  data = mpdta,
  est_method = ddml_did_method,
  est_method_vars = "fold_id",
  # --- forwarded to ddml_att via ... ---
  learners = list(what = mdl_xgboost,
                  args = list(nround = 300,
                              max_depth = 2)),
  sample_folds = K,
  silent = TRUE)
```


### Short-Stacking

Instead of committing to a single learner, we can simultaneously consider multiple learners and optimally combine them via non-negative least squares (NNLS).  Crucially, the stacking ensemble includes the linear/logistic specifications used by **did**'s default estimator, so machine learners are only selected if they improve out-of-sample fit.

*Note: small-sample and trimming warnings are suppressed below for readability.*

```{r stacking_run, eval = has_ddml, warning = FALSE}
stacking_learners <- list(
  list(what = ols, args = list(const = FALSE)),
  list(what = mdl_xgboost,
       args = list(nround = 100, max_depth = 1)),
  list(what = mdl_xgboost,
       args = list(nround = 300, max_depth = 1)),
  list(what = mdl_xgboost,
       args = list(nround = 100, max_depth = 2)),
  list(what = mdl_xgboost,
       args = list(nround = 300, max_depth = 2)))

stacking_learners_DX <- list(
  list(what = mdl_glm),
  list(what = mdl_xgboost,
       args = list(nround = 100, max_depth = 1)),
  list(what = mdl_xgboost,
       args = list(nround = 300, max_depth = 1)),
  list(what = mdl_xgboost,
       args = list(nround = 100, max_depth = 2)),
  list(what = mdl_xgboost,
       args = list(nround = 300, max_depth = 2)))

attgt_stacking <- att_gt(
  yname = "lemp",
  gname = "first.treat",
  idname = "countyreal",
  tname = "year",
  xformla = ~lpop,
  data = mpdta,
  est_method = ddml_did_method,
  est_method_vars = "fold_id",
  # --- forwarded to ddml_att via ... ---
  learners = stacking_learners,
  learners_DX = stacking_learners_DX,
  ensemble_type = "nnls",
  sample_folds = K,
  shortstack = TRUE,
  silent = TRUE)
```


## Re-estimation with a Different Ensemble

The `extra_gt` output captures the full `ddml_att` fit for each group-time cell. These stored fits contain the cross-fitted predictions (`$fitted`) and sample splits (`$splits`), which can be passed back to `ddml_att` to recompute estimates with a different ensemble type *without re-fitting any learners*.

Since `att_gt` forwards the current cell's group (`g`) and time period (`t`) to custom estimation methods, we can extend `ddml_did_method` to accept a `prev_fits` argument and look up the correct stored fit by cell identity:

```{r passthrough_wrapper, eval = has_ddml}
ddml_did_reestimate <- function(y1, y0, D, covariates, i.weights, inffunc,
                                data = NULL, g = NULL, t = NULL, ...) {
  delta_y <- y1 - y0
  dots <- list(...)

  # Look up stored fit by (g,t) key for re-estimation
  if (!is.null(dots$prev_fits) && !is.null(g) && !is.null(t)) {
    key <- paste(g, t, sep = ".")
    prev <- dots$prev_fits[[key]]
    dots$prev_fits <- NULL  # don't forward to ddml_att
    if (!is.null(prev) && !is.null(prev$ddml_fit)) {
      prev_fit <- prev$ddml_fit
      dots$learners <- prev_fit$learners
      dots$learners_DX <- prev_fit$learners_DX
      dots$fitted <- prev_fit$fitted
      dots$splits <- prev_fit$splits
    }
  }

  att_fit <- do.call(ddml_att, c(list(y = delta_y, D = D, X = covariates),
                                 dots))

  list(ATT = coef(att_fit),
       att.inf.func = att_fit$scores[, 1, 1],
       ddml_fit = att_fit)
}
```

We build a named lookup from the stacking result and re-estimate using the single-best learner ensemble, skipping all cross-fitting:

```{r passthrough_run, eval = has_ddml, warning = FALSE}
# Build named lookup: "g.t" -> extra_gt entry
prev_fits <- setNames(
  attgt_stacking$extra_gt,
  paste(attgt_stacking$group, attgt_stacking$t, sep = ".")
)

attgt_singlebest <- att_gt(
  yname = "lemp",
  gname = "first.treat",
  idname = "countyreal",
  tname = "year",
  xformla = ~lpop,
  data = mpdta,
  est_method = ddml_did_reestimate,
  # --- forwarded via ... ---
  prev_fits = prev_fits,
  ensemble_type = "singlebest",
  shortstack = TRUE,
  silent = TRUE)
```


## Comparing Results

We can compare the group-time average treatment effects across all estimators:

```{r comparison, eval = has_ddml}
comparison <- data.frame(
  group = attgt_lm$group,
  time = attgt_lm$t,
  linear = round(attgt_lm$att, 4),
  xgb_d1 = round(attgt_xgboost$att, 4),
  xgb_d2 = round(attgt_xgboost2$att, 4),
  stacking = round(attgt_stacking$att, 4),
  singlebest = round(attgt_singlebest$att, 4)
)
print(comparison)
```

The dynamic treatment effects tell a similar story:

```{r comparison_dyn, eval = has_ddml}
dyn_xgboost <- aggte(attgt_xgboost, type = "dynamic")
dyn_xgboost2 <- aggte(attgt_xgboost2, type = "dynamic")
dyn_stacking <- aggte(attgt_stacking, type = "dynamic")
dyn_singlebest <- aggte(attgt_singlebest, type = "dynamic")

dyn_comparison <- data.frame(
  event_time = dyn_lm$egt,
  linear = round(dyn_lm$att.egt, 4),
  xgb_d1 = round(dyn_xgboost$att.egt, 4),
  xgb_d2 = round(dyn_xgboost2$att.egt, 4),
  stacking = round(dyn_stacking$att.egt, 4),
  singlebest = round(dyn_singlebest$att.egt, 4)
)
print(dyn_comparison)
```

```{r stacking_dyn_plot, eval = has_ddml, fig.width=8, fig.height=5, fig.align='center', out.width="90%", dpi = 200}
ggdid(dyn_stacking, ylim = c(-.4, .4))
```


## Learner Diagnostics

The `extra_gt` output provides per-cell access to the full `ddml_att` fit objects, enabling rich learner diagnostics without global assignment. The `diagnostics()` function from `ddml` summarizes per-learner MSPE, R-squared, and ensemble weights:

```{r diagnostics, eval = has_ddml}
# Diagnostics for the first group-time cell
diagnostics(attgt_stacking$extra_gt[[1]]$ddml_fit)
```

We can also summarize stacking weights across all group-time pairs:

```{r weight_summary, eval = has_ddml}
n_gt <- length(attgt_stacking$extra_gt)

# Extract outcome and propensity score weights across all (g,t) cells
w_y_list <- lapply(seq_len(n_gt), function(i) {
  fit <- attgt_stacking$extra_gt[[i]]$ddml_fit
  if (!is.null(fit)) fit$ensemble_weights$y_X_D0 else NULL
})
w_d_list <- lapply(seq_len(n_gt), function(i) {
  fit <- attgt_stacking$extra_gt[[i]]$ddml_fit
  if (!is.null(fit)) fit$ensemble_weights$D_X else NULL
})

# Remove NULLs (skipped cells)
w_y_list <- Filter(Negate(is.null), w_y_list)
w_d_list <- Filter(Negate(is.null), w_d_list)

if (length(w_y_list) > 0) {
  w_y_mat <- do.call(rbind, lapply(w_y_list, function(w) as.vector(w)))
  w_d_mat <- do.call(rbind, lapply(w_d_list, function(w) as.vector(w)))

  learner_names <- c("OLS/Logit", "XGB (100,d1)", "XGB (300,d1)",
                     "XGB (100,d2)", "XGB (300,d2)")
  colnames(w_y_mat) <- learner_names
  colnames(w_d_mat) <- learner_names

  summarize_weights <- function(w_mat) {
    data.frame(min  = apply(w_mat, 2, min),
               mean = apply(w_mat, 2, mean),
               max  = apply(w_mat, 2, max))
  }

  cat("Outcome reduced form---stacking weights:\n")
  print(round(summarize_weights(w_y_mat), 4))

  cat("\nPropensity score---stacking weights:\n")
  print(round(summarize_weights(w_d_mat), 4))
}
```

In this application with a single covariate, we expect the parametric learners (OLS/logit) to receive most of the stacking weight---reassuring us that the default linear estimator is adequate here.

In applications with many control variables or complex interactions, one would typically observe more weight on the nonparametric learners. See Ahrens et al. (2025) for further discussion.


## References

Ahrens A, Chernozhukov V, Hansen C, Kozbur D, Schaffer M, Wiemann T (2025). "An Introduction to Double/Debiased Machine Learning." Journal of Economic Literature, forthcoming.

Callaway B, Sant'Anna P (2021). "Difference-in-differences with multiple time periods." Journal of Econometrics, 200--230.

Chang NC (2020). "Double/debiased machine learning for difference-in-difference models." The Econometrics Journal, 177--191.
